A Study of the Effectiveness of Usage Examples in REST API Documentation

### 2017-03-19 controlled_study.tex
add acknowledgement
+keywords
better positioning for screenshots
made all suggested editing changes

\ifCLASSINFOpdf
  \usepackage{listings}
  \usepackage[pdftex]{graphicx}
  \usepackage{url}
  % \usepackage{caption}
  \usepackage{subcaption}
  \usepackage{tikz,pgfplots,pgfplotstable}
  \usepgfplotslibrary{statistics}

\else
\fi
\hyphenation{op-tical net-works semi-conduc-tor}

\begin{document}
\title{How Important are Usage Examples in REST API Documentation?}

@@

\begin{abstract}
Generating and maintaining REST API documentation with usage examples can be a time consuming and expensive process for evolving APIs. Most REST API documentation tools focus on automating the documentation of the API objects, but require manual effort for capturing realistic usage examples. So, REST API developers need to know the cost vs. benefit of providing usage examples in the documentation as well as what to include in the usage examples to prioritize the documentation efforts. To this regard, we have performed a controlled study with 26 experienced software engineers and found that REST API client developers face productivity problems with using correct data types, data formats, required HTTP headers and request body when documentation lacks usage examples. By following the REST API documentation guidelines from this paper, practitioners can reduce the errors, improve success rate and satisfaction of REST API client developers.

@@

@@

\section{Introduction}
Researchers have emphasized API documentation as key factor that impacts API usability both positively and negatively. To improve API documentation, researchers have recommended incorporating usage examples in the API documentation. REST APIs are used as the predominant application integration mechanism over the Internet. Generating and maintaining REST API documentation with usage examples can be an expensive process because most API documentation tools have a lack of support for capturing realistic usage examples.

In a resource constrained environment, it is important to understand the value of usage examples on REST API usability to allocate sufficient attention and efforts towards incorporating examples in the API documentation. While it is no surprise that examples help API client developers, REST API developers need to know what to include in the examples and why. To our knowledge, there's a lack of published papers that investigated the impact of different types of content that can be included as usage examples in REST API documentation.

The documentation of REST APIs has distinctive features compared to the documentation of local APIs such as Java libraries. For example, local API documentation commonly comprises of describing the modules, classes and interfaces with their methods. On the other hand, REST API documentation needs to include information about HTTP headers, request and response body and the data representation formats such as JSON, XML, etc. In our work we have focused on understanding the impact of usage examples within the realm of the distinctive REST API features.

We have designed and performed a controlled study with experienced software engineers. Participants were divided into two groups and given the same set of API tasks to complete. One group was given the official WordPress REST API documentation and another group was given a forked version of the documentation where three usage examples were added. Using a novel technique, we have collected 539 API calls made by the participants. We have analyzed the data and observed recurring obstacles faced by the participants while performing the API tasks using the official documentation that lacks usage examples. Our contributions on this paper is as follows:

\begin{itemize}
  \item We provide a list of obstacles that REST API client developers face while performing API tasks using documentation that lacks usage examples.
  \item We provide a list of recommendations for REST developers to be used as a guideline to incorporate usage examples in the API documentation.
  \item We provide an empirical evidence that usage examples in REST API documentation helps API client developers perform API tasks with higher developer satisfaction, less time, and better success rate.
\end{itemize}

@@

To answer the aforementioned research questions, the study has the following requirements:

\begin{itemize}
  \item \textbf{R1. Representative API.} We had to choose an existing REST API that is currently used by API client developers. Selecting a mature REST API for this study reduces the possibility of an underdeveloped solution that may be found if a new or a seldom used API is selected. A familiar domain needed to be selected so that participants are able to relate to the API features without requiring upfront training. In addition to selecting the API, we had to select tasks that are related to the core features provided by the API to represent reality.
  \item \textbf{R2. Open source.} To be able to understand the impact of usage examples, we needed to select an open-source API where we can add new examples to the documentation for performing this study. For proprietary APIs, it won't be possible to retrofit usage examples on existing API documentation without separating the examples from the documentation.
  \item \textbf{R3. Time bound.} The total time spent by each participant is required to be time bound to measure productivity. As a result, the study needs to be setup such that participants are able to focus on performing the tasks minimizing any overhead.
  \item \textbf{R4. Experienced developers.} Developers with prior experience on REST APIs need to be recruited as study participants to perform the study within a limited amount of time and in a realistic setup. Furthermore, to reduce a learning bias, only participants with no prior experince of using the WordPress REST API V2 are accepted for this study.
\end{itemize}

@@

The WordPress REST API V2 has been published and maintained since May 2015. Before January 2017, the WordPress REST API V2 was distributed as a plug-in where WordPress users could optionally install the API component. The following statistics are for the plug-in install numbers between May 2015 and October 2016:

\begin{itemize}
  \item Total installs: aprox. 248K installs of the plug-in.
  \item Average daily installs: approx. 500.
\end{itemize}

Starting January 2017, the WordPress REST API V2 is no longer required to be installed as a separate plug-in since it's pre-bundled with every WordPress install. As per the code repository on GitHub, there are a total of 99 and 46 contributors that had at least one commit to the code repository behind the API and and it's documentation, respectively. These properties satisfy R1, our requirements for using a representative API.

@@

\lstset{basicstyle=\footnotesize}
\begin{lstlisting}[language=php,breaklines=true,showspaces=false,showstringspaces=false,numbers=left,xleftmargin=2em,caption={Example of self-documenting API Code},label=wordpress_code]
public function get_item_schema() {
  $schema = array(
    '$schema'    => 'http://json-schema.org/draft-04/schema#',
    'title'      => $this->post_type,
    'type'       => 'object',
    /*
     * Base properties for every Post.
     */
    'properties' => array(
        'date'            => array(
            'description' => __( "The date the object was published, in the site's timezone." ),
            'type'        => 'string',
            'format'      => 'date-time',
            'context'     => array( 'view', 'edit', 'embed' ),
        ),...
\end{lstlisting}

On Listing \ref{wordpress_code}, line 4 specifies that this is a schema definition for the API element $Post$. Then, on Line 10, it defines $date$, one of the properties of $Post$, followed by a human readable description and and type information. Then, on line 14, the context of this property is mentioned as $view, edit, embed$, meaning that this property will be returned when the $Post$ object is returned, embedded, or can be used as an input for editing.

This self-documenting feature is leveraged to generate and publish the official API documentation as an HTML website. Figure \ref{fig:schema} shows a screenshot of the published documentation for the $date$ attribute of the $Post$ API element\footnote{\url{http://v2.wp-api.org/reference/posts/}}.

@@

The participants are requested to perform a total of six tasks using the API including one practice task. All of the tasks are related to a single API element, $Post$. The tasks get progressively difficult, and all but the last task can be performed independently of each other. Participants are requested to limit the total time on the study to a maximum of one hour. Participants are encouraged to proceed to the next task when they are either satisfied about or stuck on the right answer.

@@

To verify the response of an API call, the API client developers can use HTTP response headers and/or HTTP response body. To perform the tasks, the participants are required to use one or more of the inputs I1-4. In the following paragraphs, we describe each task with it's description and the intended test scenario against the aforementioned API input and output information. For each of the tasks, the participants are required to use the same WordPress REST API and its documentation.

\textbf{T1: ListAllPostsTask}. We ask the participants to use the WordPress REST API following its documentation to get a list of the blog posts from a live WordPress site. This is the practice task, and the inputs to answer this task are pre-filled for the participants. It allows the participants to understand the tools used for this study as well as get familiarity with the Post API. The answer for this task makes use of I1 and I2.

\textbf{T2: FilterPostsByAuthorTask.} The participants are asked to use the API to filter the list of posts by an author given the author's user name. To answer this correctly, the participants are required to first make an API call to get the numeric ID of the author given the string based user name. Then, the ID needs to be used on the Post API to filter posts by the author. This task allows us to understand the impact of usage examples on API client developers when multiple API calls need to be made to perform a task using the API. Inputs I1-2 are required to complete this task successfully.

@@

To satisfy the requirement of developers with REST API experience (R3), we have used the following criteria for recruiting the study participants:

@@

A total 26 participants from sixteen different companies and six different countries (Canada, USA, Germany, Ireland, Brazil, and,
Bangladesh) were recruited through online announcements posted on Twitter, Facebook, and software developer focused mailing lists. Table \ref{table:participants} shows a summary of the experience level of the participants on each group.

@@

\subsubsection{Process}

Two pilot studies were performed to evaluate and understand an effective process for performing this study. The first pilot study involved four participants that were invited to join the first author on this paper in-person or using a video conferencing software. The study involved tasks using two APIs, the WordPress REST API V2 and the GMail REST API. Each participant was given one of the two APIs and a set of tasks to complete using the API within an hour. Participants were given an online answer form to record the answers to the tasks. The primary findings from this pilot are as follows: 1) asking participants to use an API to perform the tasks required significant overhead time for them to setup a development environment with the proper API credentials, 2) the intermediate trial attempts of using the API are potentially more valuable than the final answer as it allows us to understand API client developer information needs that may are not answered by the documentation, 3) the number of tasks for the study had to be reduced so the participants could complete the tasks within the one hour limit, and 4) for GMail API, participants used up a large portion of their time on setting up their API credentials that requires understanding of OAuth.

@@

\begin{figure*}[tb]
\begin{center}
\fbox{\includegraphics[width=\linewidth,height=2.5in,keepaspectratio]{forked.png}}
\end{center}
\caption{Screenshot of the forked WordPress REST API documentation with a usage example}
\label{fig:forked}
\end{figure*}

Based on the lessons learned from the pilot studies, we designed the actual study process as follows: the original WordPress REST API documentation was forked and a total of 3 API usage examples were added to show listing of blog posts with query parameters for filtering, a request to create a blog post and a request to update a blog post. Figure \ref{fig:original} shows a screenshot of the original API documentation related to T4 where the API client developers are provided with a reference table describing the different properties that can be used to create a $Post$ object\footnote{\url{http://v2.wp-api.org/reference/posts/}}. Figure \ref{fig:forked} shows a screenshot of the forked API documentation with a cURL based usage example. cURL is used because it's used by elsewhere in the original API documentation. In the forked API documentation, the example shows one possible API call with realistic values for the data that is described in the reference table and associated API response headers and body.

Participants were divided into two groups, G1 and G2. G1 participants were provided with a link to the official API documentation on the web-based API explorer, and G2 participants were provided with a link to the forked API documentation with usage examples. The web-based API explorer allocated more participants to G1 compared to G2 because we wanted to better understand the impact of the lack of usage examples on API client developer productivity. However, each individual participant was randomly assigned to a group by the web-based API explorer. All participants were given the same set of API tasks and were requested to limit their participation time to a maximum of one hour. No task specific time limit was imposed because we wanted participants to spend sufficient time on each task without forcing them to move the next one. Participants were allowed to access the internet and external resources alongside the provided API documentation to perform the tasks as they'd normally use on a typical work day.

@@

\subsubsection{Data Analysis}
The same data artifact is used to analyze both quantitatively and qualitatively to answer the aforementioned research question.We analyzed the data artifact qualitatively to understand the obstacles that REST API client developers face when using an API documentation that lacks usage examples. The data artifact exported from the web-based REST API explorer included a table with raw data. For each row, the table contains the following columns: API task, participant identifier, timestamp, API trial number, I1-4, response headers and body. It allowed us to observe and categorize the different input values that the participants used to perform the given API tasks. The values for I1-4 used by the participants were manually coded to group the participant responses into categories. The categories help us determine the types of information that API client developers need in the API documentation to perform API tasks successfully. The analysis started with an empty set of codes and new codes were introduced to describe scenarios that didn't fall under  codes that were already applied. The first author on this paper applied the codes on the raw data artifact and the co-authors verified the codes randomly. To resolve disagreements, the authors discussed and updated the codes accordingly. The codes were then further categorized to higher level grouping to represent common problems faced by the study participants for each group.

For each API task attempted by each participant, we annotated the artifact with one of the following labels: successful, partially successful, and unsuccessful. Task participations are marked successful when I1-4 matches the required values for performing the given task. If a participant is able to use the correct I1-4 for one of the two tasks required to complete a single task (T2), we marked it as a partially successful. Otherwise, it's marked as unsuccessful. Based on these annotations, the answers for RQ1 was computed using the following:

@@

\begin{subfigure}{0.5\textwidth}
\centering
\begin{tikzpicture}
    \begin{axis}[
        ybar,
        xtick=data,
        xticklabels from table={\groupone}{Task},
        legend style={at={(0.5,-0.15)},
        anchor=north,legend columns=-1},
        ylabel=Average Trial API Calls,
        width=2.8in
    ]
    \addplot[pattern=crosshatch dots, pattern color=gray] table [col sep=comma, x expr=\coordindex, y=AverageAPICalls] {groupone.csv};
    \addplot[draw=black] table [col sep=comma, x expr=\coordindex, y=AverageAPICalls] {grouptwo.csv};
    \legend{G1: Using Original, G2: Using Forked}
    \end{axis}

    \addplot[smooth,mark=*,black] table [col sep=comma, x expr=\coordindex, y=SuccessRate] {grouptwo_success.csv};
      \addlegendentry{G2: Using Forked}
    \end{axis}
\end{tikzpicture}
\caption{Trial vs. Successful API Calls}
\label{graph:productivity}
\end{subfigure}
\begin{subfigure}{0.4\textwidth}
\centering
\begin{tikzpicture}
    \begin{axis}[
        ybar,
        xtick=data,
        xticklabels from table={\groupone}{Task},
        legend style={at={(0.5,-0.15)},
        anchor=north,legend columns=-1},
        ylabel=Average Time Spent (minutes),
        width=2.8in
    ]
    \addplot[pattern=crosshatch dots, pattern color=gray] table [col sep=comma, x expr=\coordindex, y=Time] {groupone_time.csv};
    \addplot[draw=black] table [col sep=comma, x expr=\coordindex, y=Time] {grouptwo_time.csv};
    \legend{G1: Using Original, G2: Using Forked}
    \end{axis}

\caption{Quantitative Comparison between G1 and G2}
\label{fig:comparison}
\end{figure*}
\subsection{Qualitative Analysis Results} % (fold)

\textbf{Observation 1.} REST API client developers want usage examples along with the definition of API elements. From analyzing the participant feedback, we've observed usage examples mentioned as the most frequent and central topic of interest. 11 of the 16 G2 participants mentioned the lack of examples with HTTP headers, request and response as the primary problem with the API documentation. On the other hand, 7 out of the 10 G2 participants mentioned the examples being helpful from the forked documentation. This is also supported by our quantitative analysis as discussed later in this section.

\textbf{Observation 2.} API client developers face problem with using right data type without examples. To successfully complete T1 participants had to use an integer ID representing an $Author$ object given the author name. In the original API documentation, the following is mentioned about the author parameter: ``author: Limit result set to posts assigned to specific authors.'' The documentation doesn't mention that the data type required is the numeric author ID, not the author name. We found 10 of the 16 participants from G1 made API calls with the author name instead of the ID. P1.12 mentioned the following: \textbf{``...to find posts for the author, I had to inspect the response to see that indeed the $author\_id$ was in there''}. The forked documentation showed one example of using numeric ID. 3 of the 10 participants from G2 ran into the same error. When faced with this error, participants used trial and error to eventually complete the task.

\textbf{Observation 3.} API client developers face problem with using the right data format without examples. Tasks T3 required the use of a correct format to represent an array of numeric post IDs to be excluded using the API. The original documentation mentions the following about this API query parameter: ``exclude: Ensure result set excludes specific IDs.'' Participants attempted to solve this task using multiple different formats for specifying the array. For example: they tried with $exclude=[1,4]$, $exclude=1\&4$, $exclude=4\&\&exclude=1$, $id!=1\&id!=4$ and other alternatives before eventually finding the right format as follows $exclude=1,4$. Participant P1.6 mentioned the following: \textbf{``It was difficult to figure out whether some of the inputs needed to be arrays or just a comma-separated list.''} 7 out of the 16 participants in G1 faced this error. The forked documentation showed one example of using multiple IDs and all G2 participants could use the right format.

\textbf{Observation 4.} REST API client developers face problem with using request headers without usage examples. Tasks T5 and T6 required the use of a HTTP request header named $Content-Type$. In the original API documentation, the following was mentioned: ``The API uses JSON exclusively as the request and response format, including error responses.''. Experienced software engineers still had problem using this HTTP header as we found from the participants in G1. P1.12 mentioned the following: \textbf{``I didn't know how to specify application/json so that I could send POST data via the body instead in the URL parameters... Examples would really help - with a sample request and a sample response.''} We were surprised to see that 14 of the 15 participants that attempted T5 and T6 from G1 had the error of not specifying the $Content-Type$ header even though they have years of experience with REST APIs. In contrast, 2 out of the 10 participants from G2 failed used this header correctly as indicated in the example in the forked documentation. Moreover, 1 of the 2 G2 participants that didn't use the header first time, later used it correctly on a subsequent attempt.

@@

@@

\section{Related Work}
\label{sec:related_work}
\subsubsection{API Learning Obstacles}
  Researchers have published several papers on the topic of API learning obstacles. Robillard et al. used a mixed approach of surveys and in-person interviews with professional software engineers to understand the API learning obstacles \cite{Robillard_what_makes} \cite{Robillard_a_field_study}. They found that API documentation was related to one of the most severe obstacles for API learnabilty and recommended considering five factors for API documentation as follows: documentation of intent, code examples, matching APIs with scenarios, penetrability of the API, and format and presentation. Gias et al. surveyed 323 professional software engineers to understand the factors that fail a API documentation \cite{g_uddin}. They identified ambiguity, incompleteness, and incorrectness as the three severest problems that lead API documentations to fail to answer the information needs of developers. Duala-Ekoko et al. performed a study with twenty programmers to understand the types of questions that developers have while using an unfamiliar API \cite{Duala-Ekoko:2012:AAQ:2337223.2337255}. They categorized the API related information needs into twenty generic questions that can be used to analyze the quality of API documentation. Myers et al. performed a study of enterprise SOA API usability and found that the study participants had limited success finding the relevant API elements to perform an API task \cite{Myers_study}. Among other factors, they recommended providing code examples in the API documentation to help API client developers with sample solutions to different patterns of API tasks.

  In this paper, we focused on the usage examples part of API documentation to further understand the obstacles that API client developers face when the documentation contains relevant information about the API, it's intent, description of the API actions and elements, but lacks code examples.

>>>>

\ifCLASSINFOpdf
  \usepackage{listings}
  \usepackage[pdftex]{graphicx}
  \usepackage{url}
  % \usepackage{caption}
  \usepackage{subcaption}
  \usepackage{tikz,pgfplots,pgfplotstable}
  \usepgfplotslibrary{statistics}
\else
\fi
\hyphenation{op-tical net-works semi-conduc-tor}

\begin{document}
\title{A Study of the Effectiveness of Usage Examples in REST API Documentation}

@@

\begin{abstract}
Generating and maintaining REST API documentation with usage examples can be a time consuming and expensive process for evolving APIs. Most REST API documentation tools focus on automating the documentation of the API objects, but require manual effort for capturing realistic usage examples. Consequently, REST API developers need to know the cost vs. benefit of providing usage examples in the documentation as well as what to include in the usage examples to prioritize the documentation efforts. To this end, we have performed a controlled study with 26 experienced software engineers and found that REST API client developers face productivity problems with using correct data types, data formats, required HTTP headers and request body when documentation lacks usage examples. By following the REST API documentation guidelines from this paper, practitioners can reduce the errors, improve success rate and satisfaction of REST API client developers.

@@

\begin{IEEEkeywords}
API; REST; Documentation; Usage Examples; Empirical Study; Controlled Study; Productivity;

\end{IEEEkeywords}

@@

\section{Introduction}
Researchers have emphasized API documentation as a key factor that impacts API usability both positively and negatively. To improve API documentation, researchers have recommended incorporating usage examples in the API documentation. REST APIs are used as the predominant application integration mechanism over the Internet. Generating and maintaining REST API documentation with usage examples can be an expensive process because most API documentation tools do not support effective usage examples.

In a resource constrained environment, it is important to understand the value of usage examples on REST API usability to allocate sufficient attention and efforts towards incorporating examples in the API documentation. While it is no surprise that examples help API client developers, REST API developers need to know what to include in the examples and why.

The documentation of REST APIs has distinctive features compared to the documentation of local APIs such as Java libraries. For example, local API documentation commonly comprises the description of classes and interfaces with their methods. In contrast, REST API documentation needs to include information about HTTP headers, request and response body and the data representation formats such as JSON, XML, etc. In our work we have focused on understanding the impact of usage examples within the realm of the distinctive REST API features.

We designed and performed a controlled study with experienced software engineers. Participants were divided into two groups and given the same set of API tasks to complete. One group was given the official WordPress REST API documentation and another group was given a enhanced version of the documentation where three usage examples were added. Using a novel technique, we collected 539 API calls made by the participants. We analyzed the data and observed recurring obstacles faced by the participants while performing the API tasks using the official documentation that lacks usage examples. Our contributions on this paper is as follows:

\begin{itemize}
  \item We provide a list of obstacles that REST API client developers face while performing API tasks using documentation that lacks usage examples.
  \item We provide a list of recommendations for REST developers to be used as a guideline to incorporate usage examples in the API documentation.
  \item We provide empirical evidence that usage examples in REST API documentation help API client developers perform API tasks with higher developer satisfaction, less time, and better success rate.
\end{itemize}

@@

To answer the aforementioned research question, the study has the following requirements:

\begin{itemize}
  \item \textbf{R1. Representative API.} We had to choose an existing REST API that is currently used by API client developers. Selecting a mature REST API for this study reduces the possibility that the obstacles we observe in the study are the results of insignificant accidental problems symptomatic of an immature technology. A familiar domain needed to be selected so that participants are able to relate to the API features without requiring upfront training. In addition to selecting the API, we had to select tasks that are related to the core features provided by the API to represent reality.
  \item \textbf{R2. Open source.} To be able to understand the impact of usage examples, we needed to select an open-source API where we can add new examples to the documentation for performing this study.
  \item \textbf{R3. Time bound.} The total time spent by each participant is required to be time bound to measure productivity. As a result, the study needs to be setup such that participants are able to focus on performing the tasks minimizing any overhead.
  \item \textbf{R4. Experienced developers.} Developers with prior experience on REST APIs need to be recruited as study participants to perform the study within a limited amount of time and in a realistic setup. Furthermore, to reduce a learning bias, only participants with no prior experince of using the WordPress REST API V2 are accepted for this study.
\end{itemize}

@@

The WordPress REST API V2 has been published and maintained since May 2015. Before January 2017, the WordPress REST API V2 was distributed as a plug-in where WordPress users could optionally install the API component. The following statistics are for the plug-in installation numbers between May 2015 and October 2016:

\begin{itemize}
  \item Total installations: ~248K installs of the plug-in.
  \item Average daily installations: ~500.
\end{itemize}

Starting January 2017, the WordPress REST API V2 is no longer required to be installed as a separate plug-in since it's bundled with every WordPress installation. According to the code repository on GitHub, there are a total of 99 and 46 contributors that had at least one commit to the code repository of the API and its documentation, respectively. These properties satisfy R1, our requirements for using a representative API.

@@

\lstset{basicstyle=\footnotesize}
\begin{lstlisting}[language=php,breaklines=true,showspaces=false,showstringspaces=false,numbers=left,xleftmargin=2em,caption={Example of self-documenting API Code in PHP},label=wordpress_code]
public function get_item_schema() {
  $schema = array(
    '$schema'    => 'http://json-schema.org/draft-04/schema#',
    'title'      => $this->post_type,
    'type'       => 'object',
    /*
     * Base properties for every Post.
     */
    'properties' => array(
        'date'            => array(
            'description' => __( "The date the object was published, in the site's timezone." ),
            'type'        => 'string',
            'format'      => 'date-time',
            'context'     => array( 'view', 'edit', 'embed' ),
        ),...
\end{lstlisting}

On Listing \ref{wordpress_code}, line 4 specifies that this is a schema definition for the API element $Post$. On Line 10, it defines $date$, one of the properties of $Post$, followed by a human readable description and and type information. On line 14, the context of this property is mentioned as $view, edit, embed$, meaning that this property will be returned when the $Post$ object is returned, embedded, or can be used as an input for editing.

The WordPress team leverages the this self-documenting feature to generate and publish the official API documentation as an HTML website. Figure \ref{fig:schema} shows a screenshot of the published documentation for the $date$ attribute of the $Post$ API element.\footnote{\url{http://v2.wp-api.org/reference/posts/}}

@@

The participants are requested to perform a total of six tasks using the API including one practice task. All of the tasks are related to a single API element, $Post$ \footnote{A Post identifies a blog post within WordPress}. The tasks get progressively more difficult, and all but the last task can be performed independently of each other. Participants are requested to limit the total time on the study to a maximum of one hour. Participants are encouraged to proceed to the next task when they are either satisfied with their answer or feel stuck and unable to make progress.

@@

To verify the response of an API call, the API client developers can use HTTP response headers and/or HTTP response body. To perform the tasks, the participants are required to use one or more of the inputs I1-4. In the following paragraphs, we describe each task with its description and the study observation goals against the aforementioned API input and output information. For each of the tasks, the participants are required to use the same WordPress REST API and its documentation.

\textbf{T1: List All Posts Task}. We ask the participants to use the WordPress REST API to get a list of the blog posts from a given WordPress site. This is the practice task, and the inputs to answer this task are pre-filled for the participants. It allows the participants to understand the tools used for this study as well as get familiarity with the Post API. The answer for this task makes use of I1 and I2.

\textbf{T2: FilterPostsByAuthorTask.} The participants are asked to use the API to filter the list of posts obtained in T1 by an author given the author's user name. To answer this correctly, the participants are required to first make an API call to get the numeric ID of the author given the string based user name. Then, the ID needs to be used on the Post API to filter posts by the author. This task allows us to understand the impact of usage examples on API client developers when multiple API calls need to be made to perform a task using the API. Inputs I1-2 are required to complete this task successfully.

@@

To satisfy the requirement of developers with REST API experience (R3), we used the following criteria for recruiting the study participants:

@@

A total 26 participants from sixteen different companies and six different countries (Canada, USA, Germany, Ireland, Brazil, and,
Bangladesh) were recruited through online announcements posted on Twitter, Facebook, and software developer focused mailing lists. Table \ref{table:participants} shows a summary of the experience level of the participants in each group.

@@

\begin{figure*}[tb]
\begin{center}
\fbox{\includegraphics[width=\linewidth,height=2.5in,keepaspectratio]{forked.png}}
\end{center}
\caption{Screenshot of the enhanced WordPress REST API documentation with a usage example}
\label{fig:enhanced}
\end{figure*}

\subsubsection{Development Environment}

We conducted two pilot studies to evaluate and understand an effective process for performing this study. The first pilot study involved four participants that were invited to join the first author in-person or using a video conferencing software. The study involved tasks using two APIs, the WordPress REST API V2 and the GMail REST API. Each participant was given one of the two APIs and a set of tasks to complete using the API within an hour. We gave the participants an online answer form to record the answers to the tasks. The primary findings from this pilot are as follows: 1) asking participants to use an API to perform the tasks required significant overhead time for them to setup a development environment with the proper API credentials, 2) the intermediate trial attempts of using the API are potentially more valuable than the final answer as it allows us to understand API client developer information needs that may are not answered by the documentation, 3) the number of tasks for the study had to be reduced so the participants could complete the tasks within the one hour limit, and 4) for GMail API, participants used up a large portion of their time on setting up their API credentials that requires understanding of OAuth.

@@

\subsubsection{Protocol}

Based on the lessons learned from the pilot studies, we designed the actual study process as follows: the original WordPress REST API documentation was enhanced and a total of 3 API usage examples were added to show listing of blog posts with query parameters for filtering, a request to create a blog post and a request to update a blog post. Figure \ref{fig:original} shows a screenshot of the original API documentation related to T4 where the API client developers are provided with a reference table describing the different properties that can be used to create a $Post$ object\footnote{\url{http://v2.wp-api.org/reference/posts/}}. Figure \ref{fig:enhanced} shows a screenshot of the enhanced API documentation with a cURL based usage example. cURL is used because it's used by elsewhere in the original API documentation. In the enhanced API documentation, the example shows one possible API call with realistic values for the data that is described in the reference table and associated API response headers and body.

We divided the participants into two groups, G1 and G2. G1 participants were provided with a link to the official API documentation on the web-based API explorer, and G2 participants were provided with a link to the enhanced API documentation with usage examples. The web-based API explorer allocated more participants to G1 compared to G2 because we wanted to better understand the impact of the lack of usage examples on API client developer productivity. However, each individual participant was randomly assigned to a group by the web-based API explorer. All participants were given the same set of API tasks and were requested to limit their participation time to a maximum of one hour. No task specific time limit was imposed except an overall limit one hour for the entire study because we wanted participants to spend sufficient time on each task without forcing them to move the next one. Participants were allowed to access the internet and external resources alongside the provided API documentation to perform the tasks as they would normally use on a typical work day. We used the web-based API explorer to also collect an experience rating on a scale of 0-10, 10 being the best possible, of using the given REST API documentation and a free-form feedback from each participant as an exit survey.

@@

\begin{subfigure}{0.5\textwidth}
\centering
\begin{tikzpicture}
    \begin{axis}[
        ybar,
        xtick=data,
        xticklabels from table={\groupone}{Task},
        legend style={at={(0.5,-0.15)},
        anchor=north,legend columns=-1},
        ylabel=Average Trial API Calls,
        width=2.8in
    ]
    \addplot[pattern=crosshatch dots, pattern color=gray] table [col sep=comma, x expr=\coordindex, y=AverageAPICalls] {groupone.csv};
    \addplot[draw=black] table [col sep=comma, x expr=\coordindex, y=AverageAPICalls] {grouptwo.csv};
    \legend{G1: Using Original, G2: Using Enhanced}
    \end{axis}

    \addplot[smooth,mark=*,black] table [col sep=comma, x expr=\coordindex, y=SuccessRate] {grouptwo_success.csv};
      \addlegendentry{G2: Using Enhanced}
    \end{axis}
\end{tikzpicture}
\caption{Trial vs. Successful API Calls}
\label{graph:productivity}
\end{subfigure}
\begin{subfigure}{0.4\textwidth}
\centering
\begin{tikzpicture}
    \begin{axis}[
        ybar,
        xtick=data,
        xticklabels from table={\groupone}{Task},
        legend style={at={(0.5,-0.15)},
        anchor=north,legend columns=-1},
        ylabel=Average Time Spent (minutes),
        width=2.8in
    ]
    \addplot[pattern=crosshatch dots, pattern color=gray] table [col sep=comma, x expr=\coordindex, y=Time] {groupone_time.csv};
    \addplot[draw=black] table [col sep=comma, x expr=\coordindex, y=Time] {grouptwo_time.csv};
    \legend{G1: Using Original, G2: Using Enhanced}
    \end{axis}

\caption{Quantitative Comparison between G1 and G2}
\label{fig:comparison}
\end{figure*}

\subsubsection{Data Analysis}
We analyzed the data artifact qualitatively to understand the obstacles that REST API client developers face when using an API documentation that lacks usage examples. The data artifact exported from the web-based REST API explorer included a table with raw data. For each row, the table contains the following columns: API task, participant identifier, timestamp, API trial number, I1-4, response headers and body. It allowed us to observe and categorize the different input values that the participants used to perform the given API tasks. The values for I1-4 used by the participants were manually coded to group the participant responses into categories. The categories help us determine the types of information that API client developers need in the API documentation to perform API tasks successfully. The analysis started with an empty set of codes and new codes were introduced to describe scenarios that did not fall under  codes that were already applied. The first author applied the codes on the raw data artifact and the co-authors verified the codes randomly. To resolve disagreements, the authors discussed and updated the codes accordingly. The codes were then further categorized to higher level grouping to represent common problems faced by the study participants for each group.

For each API task attempted by each participant, we annotated the artifact with one of the following labels: successful, partially successful, and unsuccessful. Task participations are marked successful when I1-4 matches the required values for performing the given task. If a participant is able to use the correct I1-4 for one of the two tasks required to complete a single task (T2), we marked it as a partially successful. Otherwise, it's marked as unsuccessful. Based on these annotations, the answers for RQ was computed using the following:

@@

@@

\subsection{Qualitative Analysis Results} % (fold)

\textbf{Observation 1.} REST API client developers want usage examples along with the definition of API elements. From analyzing the participant feedback, we've observed usage examples mentioned as the most frequent and central topic of interest. 11 of the 16 G2 participants mentioned the lack of examples with HTTP headers, request and response as the primary problem with the API documentation. On the other hand, 7 out of the 10 G2 participants mentioned the examples being helpful from the enhanced documentation. This is also supported by our quantitative analysis as discussed later in this section.

\textbf{Observation 2.} API client developers face problem with using right data type without examples. To successfully complete T1 participants had to use an integer ID representing an $Author$ object given the author name. In the original API documentation, the following is mentioned about the author parameter: ``author: Limit result set to posts assigned to specific authors.'' The documentation doesn't mention that the data type required is the numeric author ID, not the author name. We found 10 of the 16 participants from G1 made API calls with the author name instead of the ID. P1.12 mentioned the following: \textbf{``...to find posts for the author, I had to inspect the response to see that indeed the $author\_id$ was in there''}. The enhanced documentation showed one example of using numeric ID. 3 of the 10 participants from G2 ran into the same error. When faced with this error, participants used trial and error to eventually complete the task.

\textbf{Observation 3.} API client developers face problem with using the right data format without examples. Tasks T3 required the use of a correct format to represent an array of numeric post IDs to be excluded using the API. The original documentation mentions the following about this API query parameter: ``exclude: Ensure result set excludes specific IDs.'' Participants attempted to solve this task using multiple different formats for specifying the array. For example: they tried with $exclude=[1,4]$, $exclude=1\&4$, $exclude=4\&\&exclude=1$, $id!=1\&id!=4$ and other alternatives before eventually finding the right format as follows $exclude=1,4$. Participant P1.6 mentioned the following: \textbf{``It was difficult to figure out whether some of the inputs needed to be arrays or just a comma-separated list.''} 7 out of the 16 participants in G1 faced this error. The enhanced documentation showed one example of using multiple IDs and all G2 participants could use the right format.

\textbf{Observation 4.} REST API client developers face problem with using request headers without usage examples. Tasks T5 and T6 required the use of a HTTP request header named $Content-Type$. In the original API documentation, the following was mentioned: ``The API uses JSON exclusively as the request and response format, including error responses.''. Experienced software engineers still had problem using this HTTP header as we found from the participants in G1. P1.12 mentioned the following: \textbf{``I didn't know how to specify application/json so that I could send POST data via the body instead in the URL parameters... Examples would really help - with a sample request and a sample response.''} We were surprised to see that 14 of the 15 participants that attempted T5 and T6 from G1 had the error of not specifying the $Content-Type$ header even though they have years of experience with REST APIs. In contrast, 2 out of the 10 participants from G2 failed used this header correctly as indicated in the example in the enhanced documentation. Moreover, 1 of the 2 G2 participants that didn't use the header first time, later used it correctly on a subsequent attempt.

@@

\section{Related Work}
\label{sec:related_work}
\subsubsection{API Learning Obstacles}
  Researchers have published several papers on the topic of API learning obstacles. Robillard et al. used a mixed approach of surveys and in-person interviews with professional software engineers to understand the API learning obstacles \cite{Robillard_what_makes} \cite{Robillard_a_field_study}. They found that many obstacles for API learnabilty were related to API documentation and five impactful factors for API documentation as follows: documentation of intent, code examples, matching APIs with scenarios, penetrability of the API, and format and presentation. Uddin and Robillard surveyed 323 professional software engineers to understand the different ways in which a piece of documentation can be unfit for purpose \cite{g_uddin}. They identified ambiguity, incompleteness, and incorrectness as the three severest problems that lead API documentations to fail to answer the information needs of developers. Duala-Ekoko and Robillard performed a study with twenty programmers to understand the types of questions that developers have while using an unfamiliar API \cite{Duala-Ekoko:2012:AAQ:2337223.2337255}. They categorized the API related information needs into twenty generic questions that can be used to analyze the quality of API documentation. Myers et al. performed a study of enterprise SOA API usability and found that the study participants had limited success finding the relevant API elements to perform an API task \cite{Myers_study}. Among other factors, they recommended providing code examples in the API documentation to help API client developers with sample solutions to different patterns of API tasks.

Our work fits in the general space of studies of API documentation obstacles by focusing specifically on problems related to the absence of examples, and by expanding into the study of REST APIs.
